---
title: "Formula E champion analysis"
author: "Table_5_tekkers: Aiman, Jamie, Farhan, Hylda, Joel, Yara"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load-lib, include = FALSE}
library(tidyverse)
library(tidymodels)
# Add any other libraries here



```


```{r load-data, include=FALSE}
# load your data here and any pre-processing/cleaning 
#  that you need for your report.

formula_e_race_results <- read.csv("data/formula_e_race_results.csv")



```


## Research Question

Describe your research question here. Include why this question might be of interest. 

How is driver performance affected by their team, the track, their age and any other factors?

This is interesting as it could help to make predictions, for example, about whether a driver would improve their performance by changing team. Many people follow Formula racing and this is a relevant topic, meaning our results, if communicated effectively, is likely to interest the wider population.



## Data

Briefly describe your dataset here - include a description of the key variables you will be investigating. 
Include a citation for your data. 
See http://libraryguides.vu.edu.au/c.php?g=386501&p=4347840 for guidance on proper citation for datasets. 
If you got your data off the web, make sure to state the retrieval date.



Our data set contains data from Formula E races across Seasons 1-7. There are 1,480 observations, each representing a particular driver in a particular race. There are 17 variables including the driver, the driver's team, the season number, the race number within the season and the season points awarded.

Season number is between 1 and 7, with Season 7 being incomplete as the data set was produced during this season. (Season 7 contains only 48 observations compared to 180-290 observations in the other 6 seasons)

Race number varies between 1 and 13 showing that there are up to 13 races in any given season. This is sequential meaning race number can be used to track performance from start to end of the season.

Season points shows how a driver performed during the race. A driver who scored a non-zero number of season points finished in the top 10, with more points meaning a higher finish. A season point score of 0 means the driver did not finish in the top 10 for the race.

*Citation*
M. Landry. (2021). 'Formula E Championship'. Available at https://www.kaggle.com/datasets/mlandry/formula-e-championship/data. Accessed 18/10/2024.


## Findings

The write-up of your project and findings go here. 

Think of this as the text of your presentation with some extra detail to cover what there was not time to discuss in the presentation. This might include any assumptions you made when doing your analysis, any limitations of the work you have done or any ideas you have for future work. Feel free to split this section into subsections to make it easier to read. 

The length should be roughly 1,500 words. 
If you want to use a word count addin, you can install this by copying and pasting the following into RStudio:

`devtools::install_github("benmarwick/wordcountaddin", type = "source", dependencies = TRUE)`

You will then need to restart RStudio. Once you have done that, select the text you want to count the words of, go to Addins, and select the `Word count` addin.
This addin counts words using two different algorithms, but the results should be similar and as long as you're in the ballpark of 1,500 words, you're good! 
The addin will ignore code chunks and only count the words in prose. If you don't want to use the addin you can always copy and paste the text into Microsoft Word to do a Word Count!

You can also load your data here and present any analysis results / plots.
Make sure to hide your code with  `echo = FALSE` unless the point you are trying to make is about the code itself, in which case you should show your code.





## References

M. Landry. (2021). 'Formula E Championship'. Available at https://www.kaggle.com/datasets/mlandry/formula-e-championship/data. Accessed 18/10/2024.

